Maximum Entropy Spectral Analysis: a case study

### 2021-05-27 paper/mesa.tex
Clarifications in conclusion

\section{Comparison with Welch method}\label{sec:Welch_comparison}
\sschmidt{Do we want to make a comparison with BayesLine? I think so... \\ Do we want to introduce this method in the introduction? Makes sense}
In this section we perform a qualitative comparison between the performance of the MESA and of the standard Welch algorithm.
Although similar studies can be drawn from any other PSD, for definiteness, we focus on the analytical PSD computed for LIGO Handford interferometer, released together with the GWCT-1 catalog~\citep{GWTC1, PSD_release}, and computed with BayesLine package \citep{Cornish_2015, Littenberg_2015, Chatziioannou_2019}.

@@

This observation suggests a promising avenue to pursue in future developments of gravitational waves data analysis: for short time series, comparable with the length of binary black hole systems as observed by LIGO, Virgo and KAGRA \citep{LIGO_2015, Virgo_2014, KAGRA_2012, KAGRA_2013}, the computational cost of MESA is moderate and the inferred PSD is an accurate representation of the true underlying PSD, hence MESA could be employed for a simultaneous inference of the PSD and of the gravitational wave signal during a Bayesian parameter estimation exercise, effectively marginalising over the detector noise PSD.
\sschmidt{There might be a misunderstanding: I guess the referee is talking about taking into account the PSD errors, while we are talking about dropping the stationariety assumption... The misleading word is `marginalize', I think}
This is possible whenever the log-likelihood of the model for a time series takes the form
\begin{equation}
	\log\mathcal{L}(d_t | \theta) \propto  -\frac{1}{2} \frac{(\tilde{d}(f)- \tilde{x}(f;\theta))^2}{S(f)} - \frac{1}{2}\log\left[\int  df\,S(f)\right]
\end{equation}
where $\tilde{\hphantom{a}}$ denotes the Fourier transform and the signal model $x(t;\theta)$, dependent on some parameter $\theta$, is a prediction for a deterministic signal buried in the observed time series $d(t)$.
In GW data analysis, a typical approach is to estimate the PSD offline on a large batch of data and off-source \cite{lalinference}: this scheme assumes stationary data and it might not reflect the structure of noise in the time slice to analyse.
Several approaches have been developed to try to study the validity of the assumption or even to drop it \citep{R_ver_2010, R_ver_2011, Littenberg_2013, Biscoveanu_2020, talbot2020gravitationalwave, Edwards_2020, Edwards_2015} \sschmidt{All these citations seems a bit too much for me...}.
Besides them, MESA can offer an elegant way out: at each evaluation of the likelihood, a new spectral analysis is performed by computing the PSD on $d_t-x_t$. In this way, the PSD would depend also on $\theta$ and would effectively model the residuals. Preliminary studies suggest that this is possible~\citep{martini_thesis}. Furthermore, MESA provides a simple, but robust and quite accurate, albeit for short times, predictor for the time series. This fact is remarkable and can be used in time series analysis for several purposes. As discussed in Sec.~\ref{sec:LIGO_forecasting}, an anomaly detection pipeline could be built using the forecasts of MESA: the predictions can form a baseline to compare the actual observations with. Whenever the observed data are outside the expectations, an anomaly detection can be claimed. Of course such predictions can be done with a more accurate (perhaps nonlinear) model; however MESA has the advantage of being simple and fast to construct, while providing decent predictions. At the same time, several instruments present gaps in their data stream, for instance LISA is expected to show such gaps~(e.g. \citet{lisa_gaps} and references therein), MESA forecasting capabilities could be used to fill those gaps with predicted data from past observations. In conclusion, we reiterate that MESA is a theoretically sound, computationally feasible and reliable way of studying the properties of stochastic processes and we hope that the investigations presented in this work will further stimulate developments and applications of this method.

>>>>

\section{Comparison with Welch method}\label{sec:Welch_comparison}
\sschmidt{Do we want to make a comparison with BayesLine? I think so... \\ Do we want to introduce this method in the introduction? Makes sense}
In this section we perform a qualitative comparison between the performance of the MESA and of the standard Welch algorithm.
Although similar studies can be drawn from any other PSD, for definiteness, we focus on the analytical PSD computed for LIGO Handford interferometer, released together with the GWCT-1 catalog~\citep{GWTC1, PSD_release}, and computed with BayesLine package \citep{Cornish_2015, Littenberg_2015, Cornish_2020, Chatziioannou_2019}.

@@

This observation suggests a promising avenue to pursue in future developments of gravitational waves data analysis: for short time series, comparable with the length of binary black hole systems as observed by LIGO, Virgo and KAGRA \cite{LIGO_2015, Virgo_2014, KAGRA_2012, KAGRA_2013}, the computational cost of MESA is moderate and the inferred PSD is an accurate representation of the true underlying PSD, hence MESA could be employed for online PSD estimation during a Bayesian parameter estimation exercise.
%a simultaneous inference of the PSD and of the gravitational wave signal during a Bayesian parameter estimation exercise, effectively marginalizing over the detector noise PSD.
%\sschmidt{There might be a misunderstanding: I guess the referee is talking about taking into account the PSD errors, while we are talking about dropping the stationariety assumption... The misleading word is `marginalize', I think}
This is possible whenever the log-likelihood of the model for a time series takes the form
\begin{equation}
	\log\mathcal{L}(d_t | \theta) \propto  -\frac{1}{2} \frac{(\tilde{d}(f)- \tilde{x}(f;\theta))^2}{S(f)} - \frac{1}{2}\log\left[\int  df\,S(f)\right]
\end{equation}
where $\tilde{\hphantom{a}}$ denotes the Fourier transform and the signal model $x(t;\theta)$, dependent on some parameter $\theta$, is a prediction for a deterministic signal buried in the observed time series $d(t)$.
In GW data analysis, a typical approach is to estimate the PSD offline on a large batch of data and off-source \cite{lalinference}: this scheme assumes {\it stationary data} on long timescale and it might not reflect the structure of the noise in the time slice to analyse.
Some alternatives exist for dropping this assumption and modifying the likelihood accordingly \cite{R_ver_2010, R_ver_2011, Edwards_2020, Edwards_2015}.
MESA can add to those an elegant way out: at each evaluation of the likelihood, a new spectral analysis is performed by computing the PSD on the residual $d_t-x_t$.
In this way, the PSD would depend also on $\theta$ and would effectively model the residuals. For this method, we would only need to assume the stationariety of the residual on the batch to analyse\footnote{
Unlike several works in GW data analysis \cite{Littenberg_2013, Biscoveanu_2020, talbot2020gravitationalwave}, this approach does not address the issue of including the PSD uncertainties in the posterior. Instead, it focuses on making the stationariety assumption as minimal as possible.
}, getting, as a bonus, a posterior distribution for the PSD of the analyzed data.
Preliminary studies suggest that this is possible~\cite{martini_thesis}.

%Several approaches have been developed to try to study the validity of the assumption or even to drop it \cite{R_ver_2010, R_ver_2011, Littenberg_2013, Biscoveanu_2020, talbot2020gravitationalwave, Edwards_2020, Edwards_2015} \sschmidt{All these citations seems a bit too much for me...}.
%Besides them, MESA can offer an elegant way out: at each evaluation of the likelihood, a new spectral analysis is performed by computing the PSD on $d_t-x_t$.
%In this way, the PSD would depend also on $\theta$ and would effectively model the residuals. Preliminary studies suggest that this is possible~\cite{martini_thesis}.

Furthermore, MESA provides a simple, but robust and quite accurate, albeit for short times, predictor for the time series. This fact is remarkable and can be used in time series analysis for several purposes. As discussed in Sec.~\ref{sec:LIGO_forecasting}, an anomaly detection pipeline could be built using the forecasts of MESA: the predictions can form a baseline to compare the actual observations with. Whenever the observed data are outside the expectations, an anomaly detection can be claimed. Of course such predictions can be done with a more accurate (perhaps nonlinear) model; however MESA has the advantage of being simple and fast to construct, while providing decent predictions. At the same time, several instruments present gaps in their data stream, for instance LISA is expected to show such gaps~(e.g. \citet{lisa_gaps} and references therein), MESA forecasting capabilities could be used to fill those gaps with predicted data from past observations. In conclusion, we reiterate that MESA is a theoretically sound, computationally feasible and reliable way of studying the properties of stochastic processes and we hope that the investigations presented in this work will further stimulate developments and applications of this method.
